---
title: "Displacement Power Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Displacement Power Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(ssmpower)
library(ggplot2)
```

## Introduction

Angular displacement (θ) indicates *where* on the circumplex a profile peaks. While amplitude tells us *how much* differentiation exists, displacement tells us *what kind* of differentiation—whether someone is primarily dominant, submissive, warm, cold, or some blend.

This vignette covers power analysis for displacement comparisons, based on Study 2 of Gilbert (2025).

## The Key Insight: Precision Depends on Amplitude

The most important thing to understand about displacement is that **its precision is inversely proportional to amplitude**:
$$SE(\delta) = \frac{k_\delta}{a \times \sqrt{n}}$$

where:

- $k_\delta \approx 31°$ (empirically validated scaling constant)
- $a$ = amplitude
- $n$ = sample size

This has profound implications: when amplitude is low, displacement becomes essentially undefined—there's no clear peak to locate.

```{r precision-demo}
# Show how SE changes with amplitude
amplitudes <- seq(0.10, 0.50, by = 0.05)
n <- 100

se_values <- 31 / (amplitudes * sqrt(n))

data.frame(
  Amplitude = amplitudes,
  SE_degrees = round(se_values, 1)
)
```

## When Is Displacement Analysis Appropriate?

Displacement-focused analyses are appropriate when:

1. **Amplitude is adequate** (≥ 0.15, ideally ≥ 0.20)
2. **The research question concerns angular location**, not just whether differentiation exists
3. **Prior research suggests reliable displacement** in the target construct

Displacement is **not appropriate** when:

- Expected amplitude is low (< 0.15)
- The profile is expected to be relatively flat
- You're primarily interested in whether any differentiation exists (use amplitude instead)

## Power Calculation

Use `ssm_power_displacement()` to calculate power for detecting a displacement difference:

```{r power-example}
# Power to detect 30° difference with amplitude = 0.25, n = 100 per group
ssm_power_displacement(
  delta_diff = 30,
  amplitude = 0.25,
  n = 100,
  two_group = TRUE
)
```

Compare how power changes with amplitude:

```{r power-by-amplitude}
# Same sample size, different amplitudes
amplitudes <- c(0.15, 0.20, 0.25, 0.30, 0.40)

results <- sapply(amplitudes, function(a) {
  ssm_power_displacement(delta_diff = 30, amplitude = a, n = 100)$power
})

data.frame(
 Amplitude = amplitudes,
  Power = round(results, 3)
)
```

Notice how dramatically power increases with amplitude!

## Sample Size Calculation

Use `ssm_sample_size_displacement()` to determine required sample sizes:

```{r sample-size-example}
# Sample size for 30° difference, amplitude = 0.25, 80% power
ssm_sample_size_displacement(
  delta_diff = 30,
  amplitude = 0.25,
  power = 0.80
)
```

### Sample Size Depends Heavily on Amplitude

```{r sample-size-by-amplitude}
# Required n for different amplitudes
amplitudes <- c(0.15, 0.20, 0.25, 0.30, 0.40)

results <- sapply(amplitudes, function(a) {
  ssm_sample_size_displacement(delta_diff = 30, amplitude = a)$n
})

data.frame(
  Amplitude = amplitudes,
  N_per_group = results
)
```

With low amplitude (0.15), you need nearly 4x the sample size compared to moderate amplitude (0.30).

## Precision Reference Table

The `ssm_displacement_precision_table()` function generates a quick reference:

```{r precision-table}
ssm_displacement_precision_table()
```

This shows the standard error of displacement (in degrees) for each combination of amplitude and sample size.

## Visualizing Power Curves

```{r power-curves, fig.height=6}
# Power curves for different amplitudes
ns <- seq(20, 300, by = 10)
amplitudes <- c(0.15, 0.20, 0.25, 0.30)

results <- expand.grid(n = ns, amplitude = amplitudes)
results$power <- mapply(function(n, a) {
  ssm_power_displacement(delta_diff = 30, amplitude = a, n = n)$power
}, results$n, results$amplitude)

ggplot(results, aes(x = n, y = power, color = factor(amplitude))) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "gray50") +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_color_brewer(palette = "Set1", name = "Amplitude") +
  labs(
    x = "Sample Size (per group)",
    y = "Statistical Power",
    title = "Power to Detect 30° Displacement Difference",
    subtitle = "Two-group comparison, α = .05"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Worked Example: Narcissism and Leadership Style

Suppose you want to compare the interpersonal style of narcissistic vs. non-narcissistic leaders. Prior research suggests:

- Narcissistic leaders show dominant-hostile style (θ ≈ 315°)
- Non-narcissistic leaders show dominant-warm style (θ ≈ 45°)
- Expected displacement difference: ~90° (a full quadrant)
- Expected amplitude: ~0.25 (medium)

```{r worked-example}
# How many participants per group?
result <- ssm_sample_size_displacement(
  delta_diff = 90,
  amplitude = 0.25,
  power = 0.80
)

cat("Required sample size:", result$n, "per group\n")
cat("Total N:", result$n * 2, "\n")
cat("Achieved power:", round(result$achieved_power * 100, 1), "%\n")
```

For a more conservative 45° difference:

```{r worked-example-conservative}
result <- ssm_sample_size_displacement(
  delta_diff = 45,
  amplitude = 0.25,
  power = 0.80
)

cat("Required sample size:", result$n, "per group\n")
cat("Total N:", result$n * 2, "\n")
```

## Unequal Group Sizes

For unequal allocation, use `ssm_sample_size_displacement_diff()`:

```{r unequal-groups}
# 2:1 allocation ratio
ssm_sample_size_displacement_diff(
  delta_diff = 30,
  amplitude = 0.25,
  ratio = 2
)
```

## Guidelines for Displacement-Focused Studies

1. **Verify amplitude first**: Before planning a displacement study, confirm that expected amplitude is at least 0.15, ideally 0.20+.

2. **Consider the amplitude-precision tradeoff**: If amplitude is uncertain, run sensitivity analyses across a range of plausible values.

3. **When amplitude is low, reconsider the research question**: If expected amplitude is < 0.15, displacement may not be the right focus. Consider:
   - Testing elevation differences instead
   - Testing amplitude differences
   - Using a different measure with stronger circumplex structure

4. **Report amplitude alongside displacement**: Always report amplitude estimates and CIs alongside displacement findings, so readers can evaluate precision.

## Summary

| Amplitude | Displacement Reliability | Recommendation |
|-----------|-------------------------|----------------|
| < 0.10    | Essentially undefined   | Do not analyze displacement |
| 0.10-0.15 | Highly unstable         | Caution; very large N needed |
| 0.15-0.20 | Moderately reliable     | Feasible with adequate N |
| > 0.20    | Reliable                | Standard power analysis applies |

## References

Gilbert, K. J. (2025). Powering the circumplex: A practical guide to sample size for the structural summary method.

Zimmermann, J., & Wright, A. G. C. (2017). Beyond description in interpersonal construct validation: Methodological advances in the circumplex Structural Summary Approach. *Assessment, 24*(1), 3-23.
