---
title: "Worked Example: Personality and Interpersonal Style"
author: "Kimberly Gilbert"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Worked Example: Personality and Interpersonal Style}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup, message = FALSE}
library(ssmpower)
```

## Overview

This vignette walks through a complete SSM power analysis workflow using a realistic research scenario. We'll cover:

1. A priori power analysis for study planning
2. Analyzing collected data
3. Interpreting results
4. Reporting results

## Research Scenario

**Study**: Examining how narcissistic personality traits relate to interpersonal style.

**Hypothesis**: Higher narcissism scores are associated with a distinctive interpersonal profile characterized by dominant and cold interpersonal behaviors.

**Measure**: Participants complete the Inventory of Interpersonal Problems - Short Circumplex (IIP-SC), yielding 8 octant scores:

- **PA** (0°): Domineering/Controlling
- **BC** (45°): Vindictive/Self-Centered
- **DE** (90°): Cold/Distant
- **FG** (135°): Socially Inhibited
- **HI** (180°): Nonassertive
- **JK** (225°): Overly Accommodating
- **LM** (270°): Self-Sacrificing
- **NO** (315°): Intrusive/Needy

## Part 1: A Priori Power Analysis

Before collecting data, we need to determine our required sample size.

### Step 1: Determine Expected Effect Size

Based on the literature, narcissism typically shows medium-to-large amplitude correlations with interpersonal problems (around 0.20).

```{r}
# Check what this effect size means
ssm_effect_label(0.20, "amplitude")
```

### Step 2: Calculate Required Sample Size

```{r}
# For 80% power
result_80 <- ssm_sample_size_amplitude(effect = 0.20, power = 0.80)
print(result_80)

# For 90% power (more conservative)
result_90 <- ssm_sample_size_amplitude(effect = 0.20, power = 0.90)
print(result_90)
```

### Step 3: Sensitivity Analysis

What if our effect is smaller than expected?

```{r}
cat("Sensitivity Analysis: Required N at Different Effect Sizes\n")
cat("===========================================================\n")
cat("(80% power, one-sided alpha = .05)\n\n")

effects <- c(0.15, 0.18, 0.20, 0.22, 0.25)
for (e in effects) {
  n <- ssm_sample_size_amplitude(e, power = 0.80)$n
  label <- ssm_effect_label(e, "amplitude")
  cat(sprintf("  Effect = %.2f (%s): n = %d\n", e, label, n))
}
```

### Decision

We'll aim for **N = 80** participants, which provides:
- 80% power for our expected effect (0.20)
- Adequate power (>70%) even if the true effect is 0.18

```{r}
# Verify power with N = 80
ssm_power_amplitude(effect = 0.20, n = 80)
ssm_power_amplitude(effect = 0.18, n = 80)
```

## Part 2: Data Analysis

Now we simulate data collection and analysis. (In practice, this would be real data.)

### Simulated Data

```{r}
# Simulate realistic data
set.seed(2025)
n <- 80

# Narcissism scores (standardized)
narcissism <- rnorm(n, mean = 0, sd = 1)

# IIP-SC octant scores
# Narcissism associated with high PA/BC (dominant, vindictive) and low HI/JK (assertive)
angles <- seq(0, 315, by = 45)
angles_rad <- angles * pi / 180

# Create octant scores with realistic properties
octant_scores <- matrix(NA, nrow = n, ncol = 8)
colnames(octant_scores) <- c("PA", "BC", "DE", "FG", "HI", "JK", "LM", "NO")

for (i in 1:8) {
  # Base score plus relationship with narcissism
  # Peak relationship at 45 degrees (BC - vindictive)
  loading <- 0.25 * cos(angles_rad[i] - pi/4)  # Peak at 45 degrees
  elevation_loading <- 0.15  # General interpersonal problems
  
  octant_scores[, i] <- rnorm(n, mean = 1.5, sd = 0.8) + 
                        elevation_loading * narcissism +
                        loading * narcissism
}

# Preview data
head(data.frame(narcissism = round(narcissism, 2), octant_scores[, 1:4]))
```

### Run SSM Analysis

```{r}
# Comprehensive analysis
results <- ssm_analyze(
  external_var = narcissism,
  octant_scores = octant_scores,
  n_boot = 2000,
  conf_level = 0.95
)

print(results)
```

### Interpretation

```{r}
cat("INTERPRETATION\n")
cat("==============\n\n")

cat(sprintf("1. ELEVATION: %.3f [%.3f, %.3f]\n", 
            results$elevation, 
            results$ci_elevation[1], 
            results$ci_elevation[2]))
cat(sprintf("   Effect size: %s\n", results$effect_label_elevation))
cat("   Interpretation: Higher narcissism is associated with generally\n")
cat("   elevated interpersonal problems across all octants.\n\n")

cat(sprintf("2. AMPLITUDE: %.3f [%.3f, %.3f]\n", 
            results$amplitude, 
            results$ci_amplitude[1], 
            results$ci_amplitude[2]))
cat(sprintf("   Effect size: %s\n", results$effect_label_amplitude))
cat("   Interpretation: The profile shows distinctive interpersonal\n")
cat("   differentiation - narcissism relates more to some octants than others.\n\n")

cat(sprintf("3. DISPLACEMENT: %.1f degrees [%.1f, %.1f]\n", 
            results$displacement, 
            results$ci_displacement[1], 
            results$ci_displacement[2]))
cat("   Interpretation: The peak of the profile is in the BC octant\n")
cat("   (Vindictive/Self-Centered), indicating narcissism is most strongly\n")
cat("   associated with dominant and cold interpersonal problems.\n")
```

## Part 3: Reporting Results

### For a Methods Section

```{r echo=FALSE}
cat("Sample size was determined a priori using the ssmpower package\n")
cat("(Gilbert, 2025). Based on an expected medium amplitude effect of\n")
cat("0.20, we required N = 52 for 80% power (one-sided alpha = .05).\n")
cat("We recruited N = 80 to account for potential attrition and to\n")
cat("provide adequate power for secondary analyses.\n")
```

### For a Results Section

```{r echo=FALSE}
cat(sprintf("Structural Summary Method analysis revealed that narcissism was\n"))
cat(sprintf("associated with a distinctive interpersonal profile. The profile\n"))
cat(sprintf("showed significant elevation (e = %.2f, 95%% CI [%.2f, %.2f]),\n",
            results$elevation, results$ci_elevation[1], results$ci_elevation[2]))
cat(sprintf("indicating that higher narcissism was associated with greater\n"))
cat(sprintf("overall interpersonal problems. The profile amplitude was %.2f\n",
            results$amplitude))
cat(sprintf("(95%% CI [%.2f, %.2f]), representing a %s effect and indicating\n",
            results$ci_amplitude[1], results$ci_amplitude[2],
            tolower(results$effect_label_amplitude)))
cat(sprintf("interpersonal differentiation. The profile peaked at %.0f degrees\n",
            results$displacement))
cat(sprintf("(95%% CI [%.0f, %.0f]), in the BC (Vindictive/Self-Centered) octant.\n",
            results$ci_displacement[1], results$ci_displacement[2]))
```

### Reporting Table

```{r}
# Create results table
results_table <- data.frame(
  Parameter = c("Elevation", "Amplitude", "Displacement"),
  Estimate = c(
    sprintf("%.3f", results$elevation),
    sprintf("%.3f", results$amplitude),
    sprintf("%.1f", results$displacement)
  ),
  CI_95 = c(
    sprintf("[%.3f, %.3f]", results$ci_elevation[1], results$ci_elevation[2]),
    sprintf("[%.3f, %.3f]", results$ci_amplitude[1], results$ci_amplitude[2]),
    sprintf("[%.1f, %.1f]", results$ci_displacement[1], results$ci_displacement[2])
  ),
  Effect_Size = c(
    results$effect_label_elevation,
    results$effect_label_amplitude,
    "N/A"
  )
)

knitr::kable(results_table, 
             col.names = c("Parameter", "Estimate", "95% CI", "Effect Size"),
             caption = "SSM Profile of Narcissism and Interpersonal Problems")
```

## Summary Workflow

1. **Before data collection**:
   - Use `ssm_sample_size_amplitude()` to determine required N
   - Conduct sensitivity analysis for range of plausible effects

2. **After data collection**:
   - Use `ssm_analyze()` for comprehensive analysis
   - Bootstrap CIs provide inference for all parameters

3. **Reporting**:
   - Report a priori power analysis in Methods
   - Report SSM parameters with CIs and effect size labels in Results

## Citation

To cite this package:

Gilbert, K. (2025). *ssmpower: Power Analysis for the Structural Summary Method*. R package version 1.0.0. https://github.com/kimberlyjg/ssmpower
